Pyro&thinsp;-Velocity: Probabilistic RNA Velocity inference from single-cell data

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): enable toc
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: "Pyro&thinsp;-Velocity: Probabilistic RNA Velocity inference from single-cell data"
format:
  nature-pdf:
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-nature
    classoption: 
      - lineno
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: citeproc
    keep-tex: true
    equal-margins: false
    include-in-header: header.tex
  html:
    code-fold: true
    html-math-method: mathjax
  docx:
    reference-doc: styles.docx
# filters:
#   - authors-block
csl: _extensions/sciexp/nature/csl/springer-basic-brackets-no-et-al.csl
affiliations:
  - name: Massachusetts General Hospital Research Institute, Charlestown, MA
    id: 1
    department: Molecular Pathology Unit
    city: Charlestown
    state: MA
    postal-code: 02129
  - name: Harvard Medical School, Charlestown, MA
    id: 2
    department: Massachusetts General Hospital Cancer Center
    city: Charlestown
    state: MA
    postal-code: 02129
  - name: Massachusetts General Hospital, Boston, MA
    id: 3
    department: Center for Regenerative Medicine
    city: Boston
    state: MA
    postal-code: 02114
  - name: Harvard University, Cambridge, MA
    id: 4
    department: Harvard Stem Cell Institute
    city: Cambridge
    state: MA
    postal-code: 02139
  - name: Cambridge, MA
    id: 5
    department: Broad Institute of MIT and Harvard
    city: Cambridge
    state: MA
    postal-code: 02142
  - name: Laboratory of Neurodevelopmental Systems Biology, Lausanne, Switzerland
    id: 6
    department: Brain and Mind Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne (EPFL)
    city: Lausanne
    state: Switzerland
    postal-code: 1015
  - name: Cambridge, MA
    id: 7
    department: Basis Research Institute
    city: Cambridge
    state: MA
    postal-code: 02142
authors:
  - name: Qian Qin
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
      - ref: 4
      - ref: 5
    # attributes:
    #   equal-contributor: true
  - name: Eli Bingham
    affiliations:
      - ref: 5
      - ref: 7
  - name: Gioele La Manno
    affiliations:
      - ref: 6
    email: gioele.lamanno@epfl.ch
    attributes:
      corresponding: true
  - name: David M. Langenau
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
      - ref: 4
    email: dlangenau@mgh.harvard.edu
    attributes:
      corresponding: true
  - name: Luca Pinello
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 5
    email: lpinello@mgh.harvard.edu
    attributes:
      corresponding: true
abstract: |
  Single-cell RNA Velocity has dramatically advanced our ability to model cellular differentiation and cell fate decisions. However, current preprocessing choices and model assumptions often lead to errors in assigning developmental trajectories. Here, we develop, Pyro-Velocity, a Bayesian, generative, and multivariate RNA Velocity model to estimate the uncertainty of cell future states. This approach models raw sequencing counts with the synchronized cell time across all expressed genes to provide quantifiable and improved information on cell fate choices and developmental trajectory dynamics.
keywords: [RNA velocity, single-cell RNA sequencing, probabilistic generative modeling, cell fate, developmental trajectory inference]
bibliography: bibliography.bib
crossref:
  fig-prefix: Fig.
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Fig. S
      space-before-numbering: false
      latex-list-of-description: Supplementary Figure
---

\newpage

@@

@@

::: {#fig-model}

![](figures/fig.pdf)

\label{fig-model-inputs} 
\label{fig-model-outputs}

>>>>

---
title: "Pyro&thinsp;-Velocity: Probabilistic RNA Velocity inference from single-cell data"
number-sections: true
toc: true
toc-title: Contents
format:
  nature-pdf:
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-nature
    classoption: 
      - lineno
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: citeproc
    keep-tex: true
    equal-margins: false
    include-in-header: header.tex
  html:
    code-fold: true
    html-math-method: mathjax
  docx:
    reference-doc: styles.docx
# filters:
#   - authors-block
csl: _extensions/sciexp/nature/csl/springer-basic-brackets-no-et-al.csl
affiliations:
  - name: Massachusetts General Hospital Research Institute, Charlestown, MA
    id: 1
    department: Molecular Pathology Unit
    city: Charlestown
    state: MA
    postal-code: 02129
  - name: Harvard Medical School, Charlestown, MA
    id: 2
    department: Massachusetts General Hospital Cancer Center
    city: Charlestown
    state: MA
    postal-code: 02129
  - name: Massachusetts General Hospital, Boston, MA
    id: 3
    department: Center for Regenerative Medicine
    city: Boston
    state: MA
    postal-code: 02114
  - name: Harvard University, Cambridge, MA
    id: 4
    department: Harvard Stem Cell Institute
    city: Cambridge
    state: MA
    postal-code: 02139
  - name: Cambridge, MA
    id: 5
    department: Broad Institute of MIT and Harvard
    city: Cambridge
    state: MA
    postal-code: 02142
  - name: Laboratory of Neurodevelopmental Systems Biology, Lausanne, Switzerland
    id: 6
    department: Brain and Mind Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne (EPFL)
    city: Lausanne
    state: Switzerland
    postal-code: 1015
  - name: Cambridge, MA
    id: 7
    department: Basis Research Institute
    city: Cambridge
    state: MA
    postal-code: 02142
authors:
  - name: Qian Qin
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
      - ref: 4
      - ref: 5
    # attributes:
    #   equal-contributor: true
  - name: Eli Bingham
    affiliations:
      - ref: 5
      - ref: 7
  - name: Gioele La Manno
    affiliations:
      - ref: 6
    email: gioele.lamanno@epfl.ch
    attributes:
      corresponding: true
  - name: David M. Langenau
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
      - ref: 4
    email: dlangenau@mgh.harvard.edu
    attributes:
      corresponding: true
  - name: Luca Pinello
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 5
    email: lpinello@mgh.harvard.edu
    attributes:
      corresponding: true
abstract: |
  Single-cell RNA Velocity has dramatically advanced our ability to model cellular differentiation and cell fate decisions. However, current preprocessing choices and model assumptions often lead to errors in assigning developmental trajectories. Here, we develop, Pyro-Velocity, a Bayesian, generative, and multivariate RNA Velocity model to estimate the uncertainty of cell future states. This approach models raw sequencing counts with the synchronized cell time across all expressed genes to provide quantifiable and improved information on cell fate choices and developmental trajectory dynamics.
keywords: [RNA velocity, single-cell RNA sequencing, probabilistic generative modeling, cell fate, developmental trajectory inference]
bibliography: bibliography.bib
crossref:
  fig-prefix: Fig.
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Fig. S
      space-before-numbering: false
      latex-list-of-description: Supplementary Figure
---

@@

Pyro-Velocity recasts the velocity estimation problem into a latent variable posterior probability inference (Methods) [@sec-methods]. The proposed model is generative and fully Bayesian, with the different parameters considered as latent random variables. Each parameter is assigned a prior distribution, and the model is then conditioned on the raw data (both spliced and unspliced molecule raw counts) to estimate posterior probability distributions and to enable uncertainty estimation of velocity. Central to the Pyro-Velocity model is a shared latent time (t): a global latent variable placing each cell at a particular instant of the modeled dynamical process. From this latent time, the expectation of spliced and unspliced RNA levels for each gene can be computed knowing gene-specific kinetics parameters (transcriptional (), splicing (), and degradation rates () and two gene-specific switching times (t0k), and gene-specific transcriptional state k) and integrating the RNA velocity differential equations analogously to what proposed in the dynamical RNA velocity model by Bergen et al.4,9. However, we innovate the structure and formulation of the latent time given that our model is set to jointly fit the entire raw data matrix, not individual genes. We approach the task of estimating the posterior distribution by restricting the search within a family of a variational distribution approximating the true posterior, and taking advantage of the automatic differentiation variational inference (ADVI) capabilities of  Pyro8,10 (Fig. 1a, Methods). Notably, our generative Bayesian model enables to sample from the posterior distributions of the latent and dependent variables, allowing uncertainty estimation of phase portraits, high dimensional velocity vector field, and shared latent time. Around this model, we built several visualizations to assess uncertainty and functions to recover putative cell fate priming genes and to investigate how the denoised gene expression recapitulates known cell-type annotations and their developmental order (Fig. 1b, Methods). Although several RNA velocity methods11–15 have been recently proposed, and many of them have not been peer-reviewed, here we showcase the advantages and innovations of Pyro-Velocity by comparing it with the current state-of-the-art dynamical RNA velocity model proposed by Bergen et al.4, and implemented in scVelo.

# Methods {#sec-methods}

## Model formulation {#sec-model}

We assume the dynamical gene expression is determined by the RNA splicing process, and infer the unspliced and spliced gene expression level from the differential equations proposed in velocyto [@La_Manno2018-lj] and scVelo [@Bergen2020-pj]

\begin{align}
\frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}} &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right), \label{eq-dudt}\\
\frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}} &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)-\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
\end{align}

@@

::: {#fig-model}
![](figures/fig.pdf)
